Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2023-10-16 supplementary-materials.tex
update supplementary materials to work with current gratia

@@

@@

  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}

@@

\usepackage{microtype}

@@

\hypersetup{unicode=true,

@@

\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}

@@

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Gavin L. Simpson}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{August 16, 2018}

@@

\usepackage{lineno}
\linenumbers

@@

\section{Introduction}\label{introduction}

@@

The following packages are required: \emph{mgcv}, \emph{scam},
\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}.

@@

\begin{verbatim}
#> This is mgcv 1.8-24. For overview type 'help("mgcv-package")'.
\end{verbatim}

@@

\begin{verbatim}
#> This is scam 1.2-2.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"ggplot2"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"cowplot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Attaching package: 'cowplot'
\end{verbatim}

\begin{verbatim}
#> The following object is masked from 'package:ggplot2':
#> 
#>     ggsave
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

In addition, the \emph{gratia} package is required; this in-development
package is not on CRAN but can be installed directly from GitHub using
functions from the \emph{devtools} package. To install the package,
install \emph{devtools} and then use
\texttt{devtools::install\_github()} to install \emph{gratia}, as shown
below:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## gratia is not on CRAN, install from github:}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/gratia"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default ggplot theme}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

\section{Load the data sets}\label{load-the-data-sets}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## source Small Water data}
\NormalTok{small <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"./data/small-water/small-water-isotope-data.rds"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(small)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## load braya so data set}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                    \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
\NormalTok{## clean up variable names}
\KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                  \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
\NormalTok{## add a variable for the amount of time per sediment sample}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
\KeywordTok{head}\NormalTok{(braya)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>   Depth DepthUpper DepthLower     Year YearYoung YearOld   UK37
#> 1  0.25        0.0        0.5 1999.125   2006.00 1992.25 -0.640
#> 2  0.75        0.5        1.0 1985.375   1992.25 1978.50 -0.637
#> 3  1.25        1.0        1.5 1971.525   1978.50 1964.55 -0.614
#> 4  1.75        1.5        2.0 1957.575   1964.55 1950.60 -0.627
#> 5  2.25        2.0        2.5 1943.150   1950.60 1935.70 -0.633
#> 6  2.75        2.5        3.0 1928.250   1935.70 1920.80 -0.616
#>   sampleInterval
#> 1          13.75
#> 2          13.75
#> 3          13.95
#> 4          13.95
#> 5          14.90
#> 6          14.90
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot labels}
\NormalTok{d15n_label <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(delta^\{}\DecValTok{15}\NormalTok{\}*N)}
\NormalTok{braya_ylabel <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(U)[}\DecValTok{37}\NormalTok{]^\{}\KeywordTok{italic}\NormalTok{(k)\})}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot Small Water data}
\NormalTok{small_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N)) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\NormalTok{## plot Braya-So data}
\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{UK37)) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\NormalTok{## Recreate plot from manuscript}
\KeywordTok{plot_grid}\NormalTok{(small_plt, braya_plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
          \DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\section{Fitting GAMs}\label{fitting-gams}

@@

\subsection{Small Water}\label{small-water}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## fit small water GAM using gamm() with a CAR(1)}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## estimate of phi and confidence interval}
\NormalTok{smallPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(mod$lme, }\DataTypeTok{which =} \StringTok{"var-cov"}\NormalTok{)$corStruct}
\NormalTok{smallPhi}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>        lower      est.     upper
#> Phi 0.282388 0.6026967 0.8539689
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## summary object}
\KeywordTok{summary}\NormalTok{(mod$gam)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot CAR(1) process}
\NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
                 \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =} \NormalTok{S)}

\NormalTok{car1Plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(car1, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{S, }\DataTypeTok{y =} \NormalTok{Correlation)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymax =} \NormalTok{Upper, }\DataTypeTok{ymin =} \NormalTok{Lower),}
                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(h) *}\StringTok{ }\NormalTok{(}\KeywordTok{list}\NormalTok{(Delta[t], varphi)))) +}
\StringTok{    }\KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(Delta[t] ~}\StringTok{ }\NormalTok{(years)))}
\NormalTok{car1Plt}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Braya-Sø}\label{braya-s}

The same model as used for Small Water was also initially attempted with
the Braya-Sø data. However, in order to even fit the model with both a
smooth and the CAR(1) process, I had to change the default optimizer
used to estimate the model parameters, and reduce the basis dimension,
\texttt{k}, to a small number. I also fit the model using GCV, which is,
at the time of writing, the default in \texttt{gam()}, hence no
\texttt{method} argument

>>>>

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%

@@

  \usepackage{textcomp} % provides euro and other symbols

@@

  \usepackage{unicode-math}

@@

\usepackage[]{microtype}

@@

\hypersetup{

@@

\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

%\usepackage{lineno}
%\linenumbers

\title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
\author{Gavin L. Simpson}
\date{October 16, 2023}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

The following packages are required: \emph{mgcv}, \emph{scam},
\emph{ggplot2}, \emph{cowplot}, \emph{dplyr}, and \emph{tidyr}.

@@

\begin{verbatim}
#> This is mgcv 1.9-0. For overview type 'help("mgcv-package")'.
\end{verbatim}

@@

\begin{verbatim}
#> This is scam 1.2-14.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"ggplot2"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"cowplot"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"dplyr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Attaching package: 'dplyr'
\end{verbatim}

\begin{verbatim}
#> The following object is masked from 'package:nlme':
#> 
#>     collapse
\end{verbatim}

\begin{verbatim}
#> The following objects are masked from 'package:stats':
#> 
#>     filter, lag
\end{verbatim}

\begin{verbatim}
#> The following objects are masked from 'package:base':
#> 
#>     intersect, setdiff, setequal, union
\end{verbatim}

In addition, the \emph{gratia} package is required; install
\emph{gratia}, using:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## requires development version of gratia that is not on CRAN, install from github:}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"gratia"}\NormalTok{, }\DataTypeTok{repos =} \KeywordTok{c}\NormalTok{(}\StringTok{"https://gavinsimpson.r-universe.dev"}\NormalTok{,}
    \StringTok{"https://cloud.r-project.org"}\NormalTok{))}
\KeywordTok{stopifnot}\NormalTok{(}\KeywordTok{packageVersion}\NormalTok{(}\StringTok{"gratia"}\NormalTok{) }\OperatorTok{>=}\StringTok{ "0.8.1.42"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Default ggplot theme}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

\hypertarget{load-the-data-sets}{%
\section{Load the data sets}\label{load-the-data-sets}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## source Small Water data}
\NormalTok{small <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"./data/small-water/small-water-isotope-data.rds"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(small)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## load braya so data set}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                    \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
\CommentTok{## clean up variable names}
\KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                  \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
\CommentTok{## add a variable for the amount of time per sediment sample}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =}\NormalTok{ YearYoung }\OperatorTok{-}\StringTok{ }\NormalTok{YearOld)}
\KeywordTok{head}\NormalTok{(braya)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>   Depth DepthUpper DepthLower     Year YearYoung YearOld   UK37 sampleInterval
#> 1  0.25        0.0        0.5 1999.125   2006.00 1992.25 -0.640          13.75
#> 2  0.75        0.5        1.0 1985.375   1992.25 1978.50 -0.637          13.75
#> 3  1.25        1.0        1.5 1971.525   1978.50 1964.55 -0.614          13.95
#> 4  1.75        1.5        2.0 1957.575   1964.55 1950.60 -0.627          13.95
#> 5  2.25        2.0        2.5 1943.150   1950.60 1935.70 -0.633          14.90
#> 6  2.75        2.5        3.0 1928.250   1935.70 1920.80 -0.616          14.90
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## plot labels}
\NormalTok{d15n_label <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(delta}\OperatorTok{^}\NormalTok{\{}\DecValTok{15}\NormalTok{\}}\OperatorTok{*}\NormalTok{N)}
\NormalTok{braya_ylabel <-}\StringTok{ }\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(U)[}\DecValTok{37}\NormalTok{]}\OperatorTok{^}\NormalTok{\{}\KeywordTok{italic}\NormalTok{(k)\})}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## plot Small Water data}
\NormalTok{small_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(small, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ d15N)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\CommentTok{## plot Braya-So data}
\NormalTok{braya_plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(braya, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Year, }\DataTypeTok{y =}\NormalTok{ UK37)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{colour =} \StringTok{"grey"}\NormalTok{) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ braya_ylabel, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}

\CommentTok{## Recreate plot from manuscript}
\KeywordTok{plot_grid}\NormalTok{(small_plt, braya_plt, }\DataTypeTok{ncol =} \DecValTok{1}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"auto"}\NormalTok{, }\DataTypeTok{align =} \StringTok{"hv"}\NormalTok{,}
          \DataTypeTok{axis =} \StringTok{"lr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{fitting-gams}{%
\section{Fitting GAMs}\label{fitting-gams}}

@@

\hypertarget{small-water}{%
\subsection{Small Water}\label{small-water}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## fit small water GAM using gamm() with a CAR(1)}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ small,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## estimate of phi and confidence interval}
\NormalTok{smallPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(mod}\OperatorTok{$}\NormalTok{lme, }\DataTypeTok{which =} \StringTok{"var-cov"}\NormalTok{)}\OperatorTok{$}\NormalTok{corStruct}
\NormalTok{smallPhi}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>         lower      est.     upper
#> Phi 0.2811374 0.6026967 0.8547379
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## summary object}
\KeywordTok{summary}\NormalTok{(mod}\OperatorTok{$}\NormalTok{gam)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## plot CAR(1) process}
\NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
                 \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =}\NormalTok{ S)}

\NormalTok{car1Plt <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(car1, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ S, }\DataTypeTok{y =}\NormalTok{ Correlation)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymax =}\NormalTok{ Upper, }\DataTypeTok{ymin =}\NormalTok{ Lower),}
                \DataTypeTok{fill =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{) }\OperatorTok{+}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
\StringTok{    }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(c) }\OperatorTok{*}\StringTok{ }\NormalTok{(}\KeywordTok{list}\NormalTok{(h, varphi)))) }\OperatorTok{+}
\StringTok{    }\KeywordTok{xlab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(h }\OperatorTok{~}\StringTok{ }\NormalTok{(years)))}
\NormalTok{car1Plt}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{braya-suxf8}{%
\subsection{Braya-Sø}\label{braya-suxf8}}

The same model as used for Small Water was also initially attempted with
the Braya-Sø data. I also fit the model using GCV, which is, at the time
of writing, the default in \texttt{gam()}, hence no \texttt{method}
argument

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## fit the car(1) model --- needs optim as this is not a stable fit!}
\CommentTok{## also needs k setting lower than default}
\NormalTok{braya.car1 <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{10}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya,}
                   \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \OperatorTok{~}\StringTok{ }\NormalTok{Year),}
                   \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}

\CommentTok{## fit model using GCV}
\NormalTok{braya.gcv <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 }\OperatorTok{~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{30}\NormalTok{), }\DataTypeTok{data =}\NormalTok{ braya)}

\CommentTok{## estimate of phi and confidence interval}
\NormalTok{brayaPhi <-}\StringTok{ }\KeywordTok{intervals}\NormalTok{(braya.car1}\OperatorTok{$}\NormalTok{lme)}\OperatorTok{$}\NormalTok{corStruct}
\NormalTok{brayaPhi}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>            lower      est. upper
#> Phi 1.239228e-18 0.2000199     1
#> attr(,"label")
#> [1] "Correlation structure:"
\end{verbatim}
